A new method to explicitly estimate the shift of optimum in multispecies studies

### 2022-01-24 redaction/manuscript.Rmd
Take into account TC corrections (made in manuscript+TC_29dec2021.pdf)

2) We extended an existing model to explicitly estimate optimum shifts for multiple species. We called this new model the Explicit Hierarchical Model of Optimum Shifts (EHMOS). In a simulation study, we compared its accuracy to a mean comparison method and a generalized linear mixed model (GLMM). Specifically, we tested if the accuracy of the methods was sensitive to (1) sampling design, (2) species optimum position, and (3) species ecological specialization. In addition, we compared the three methods using a real dataset of investigated optimum shifts in 24 Orthopteran species between two surveys conducted thirty years apart along an elevation gradient in the French Alps.

@@

Understanding changes in relationships between species and environmental, geographical or temporal gradients is of central interest in ecology and evolution. For instance, assessing changes in species distribution along latitudinal [@thorson2016; @ortega2019], elevational [@maggini2011] or climatic [@tayleur2015] gradients in response to ongoing climate change is crucial for developing relevant biodiversity change scenarios [@thorson2016; @ortega2019; @maggini2011]. Modelling these changes is also central in the study of invasive species to predict the area of potential colonization [@elith2010]. The shifts in species phenology, i.e. changes along a temporal gradient, in response to climate change are also well-studied to obtain a mechanistic understanding of climate change impacts on species populations [@moussus2010; @vitasse2021; @strebel2014]. Shifts that could occur along nutritional gradients during a species' lifespan or two occupied regions are also studied in a more theoretical context [@coudun2005; @bertrand2011].

@@

Motivated by a case study on Orthoptera communities sampled twice 30 years apart in the French Alps along an elevation gradient, we tested three methods for estimating optimum shifts of the entire species community. We developed an extension of the Gaussian logistic model [@jamil2014; @terbraak1986] that explicitly estimates shifts in optima for several species simultaneously, with their uncertainties. We also fitted a Bayesian Generalized Linear Mixed Model to simultaneously analyse both sampling occasions and all species and use the posterior distributions of the estimated coefficients to compute posterior distributions of species optimum shifts. The performance of these models was evaluated and compared to the estimates provided by a simple mean comparison method. The comparison was undertaken through a simulation study involving different sampling designs and ecological scenarios. The primary goal of the simulation study was to determine if the tested methods accurately estimate the optimum shifts of multiple species along a virtual environmental gradient. Specifically, we tested if the accuracy of the methods was sensitive to (1) sampling design, i.e. how sampling sites are distributed along the gradient, (2) species marginality, i.e. the location of the species optimum on the gradient, and (3) species ecological specialization, i.e. the width of the species–environment relationship.

@@

*Explicit Hierarchical Model of Optimum Shifts (EHMOS)* – We extended the multi-species Gaussian logistic model developed to describe species unique unimodal response curves [@jamil2014] to explicitly estimate shift in optimum positions between two sampling occasions. Our model, named Explicit Hierarchical Model of Optimum Shifts (EHMOS), describes multiple species–environment relationships simultaneously for two samples. Species response curves are described through four ecologically meaningful parameters ($\alpha_{i,k}$, $\theta_i$, $\delta$ and $\tau_{i,k}$). The occupancy model was written as follows:

@@

Our goal when starting this study was to find an appropriate modelling approach that produces estimates, and associated uncertainties, of shifts in elevation optima of an Orthoptera community studied during two periods separated by 30 years. The study we selected for the data was conducted in 134 sampling sites distributed along an elevation gradient ranging from 928 m to 2614 m (mean = 1869 m) above sea level in Mercantour National Park in the southern French Alps. The first survey was carried out during the summers between 1983 and 1988 (Gueguen 1990). At each site, the density of Orthoptera species was assessed based on multiple counts made using box quadrats that trapped individuals. The second survey was conducted during the summers of 2018 and 2019. Each sampling site consisted of a circle with a 70-m radius where five spatial replicates (hereafter, plots) 30 m2 in area were surveyed following three successive steps: (1) one minute of listening to species stridulating in the plot by standing close to its edge, (2) six minutes of sighting species by walking across the entire plot, and (3) two 45-second sweep netting sessions across the entire plot (see Mourguiart et al. 2021 for a full description of the sampling design). To minimize the effect of varying sampling effort and detection methods between surveys, we reduced the Orthoptera records to the presence/absence data at the site level and only kept species detected in at least 5\% of sampling sites, resulting in 24 species. We pooled some species' data in a species complex when species confusion was possible. Thus, we aggregated data for species in the genuses *Yersinella*, *Calliptamus*, *Podisma* and in the complex *Chorthippus biggutulus-brunneus-mollis-daimei*.

@@

## Data availability
The data that support the findings of this study are available from the corresponding author upon reasonable request.

>>>>

2) We extended an existing model to explicitly estimate optimum shifts for multiple species. We called this new Bayesian hierarchical model the Explicit Hierarchical Model of Optimum Shifts (EHMOS). In a simulation study, we compared its accuracy to a mean comparison method and a Bayesian generalized linear mixed model (GLMM). Specifically, we tested if the accuracy of the methods was sensitive to (1) sampling design, (2) species optimum position, and (3) species ecological specialization. In addition, we compared the three methods using a real dataset of investigated optimum shifts in 24 Orthopteran species between two surveys conducted thirty years apart along an elevation gradient in the French Alps.

@@

Understanding changes in relationships between species and environmental, geographical or temporal gradients is of central interest in ecology and evolution. For instance, assessing changes in species distribution along latitudinal [@thorson2016; @ortega2019], elevational [@maggini2011] or climatic [@tayleur2015] gradients in response to ongoing climate change is crucial for developing relevant biodiversity change scenarios [@thorson2016; @ortega2019; @maggini2011]. Modelling these changes is also central in the study of invasive species to predict the area of potential colonization [@elith2010]. The shifts in species phenology, i.e. changes along a temporal gradient, in response to climate change are also well-studied to obtain a mechanistic understanding of climate change impacts on species populations [@moussus2010; @vitasse2021; @strebel2014]. The range of topics interested in shifts also include more theoretical studies, such as research on shifts of plant nutritional niche during a species' lifespan [@bertrand2011].

@@

We developed an extension of the Gaussian logistic model [@jamil2014; @terbraak1986] that explicitly estimates shifts in optima for several species simultaneously, with their uncertainties. We also fitted a Bayesian Generalized Linear Mixed Model to simultaneously analyse both sampling occasions and all species and use the posterior distributions of the estimated coefficients to compute posterior distributions of species optimum shifts. The performance of these models was evaluated and compared to the estimates provided by a simple mean comparison method. The comparison was undertaken through a simulation study involving different sampling designs and ecological scenarios. The primary goal of the simulation study was to determine if the tested methods accurately estimate the optimum shifts of multiple species along a virtual environmental gradient. Specifically, we tested if the accuracy of the methods was sensitive to (1) sampling design, i.e. how sampling sites are distributed along the gradient, (2) species marginality, i.e. the location of the species optimum on the gradient, and (3) species ecological specialization, i.e. the width of the species–environment relationship. We finally applied the three methods to 24 Orthoptera species sampled twice 30 years apart in the French Alps along an elevation gradient.

@@

*Explicit Hierarchical Model of Optimum Shifts (EHMOS)* – We extended the multi-species Gaussian logistic model developed to describe species unique unimodal response curves [@jamil2014] to explicitly estimate shift in optimum positions between two sampling occasions. This model, we called Explicit Hierarchical Model of Optimum Shifts (EHMOS), describes multiple species–environment relationships simultaneously for two samples. Species response curves are described through four ecologically meaningful parameters ($\alpha_{i,k}$, $\theta_i$, $\delta$ and $\tau_{i,k}$). The occupancy model was written as follows:

@@

Our goal when starting this study was to find an appropriate modelling approach that produces estimates, and associated uncertainties, of shifts in elevation optima of an Orthoptera community studied during two periods separated by 30 years. This study was conducted in 134 sampling sites distributed along an elevation gradient ranging from 928 m to 2614 m (mean = 1869 m) above sea level in Mercantour National Park in the southern French Alps. The first survey was carried out during the summers between 1983 and 1988 (Gueguen 1990). At each site, the density of Orthoptera species was assessed based on multiple counts made using box quadrats that trapped individuals. The second survey was conducted during the summers of 2018 and 2019. Each sampling site consisted of a circle with a 70-m radius where five spatial replicates (hereafter, plots) 30 m2 in area were surveyed following three successive steps: (1) one minute of listening to species stridulating in the plot by standing close to its edge, (2) six minutes of sighting species by walking across the entire plot, and (3) two 45-second sweep netting sessions across the entire plot [see @mourguiart2021 for a full description of the sampling design]. To minimize the effect of varying sampling effort and detection methods between surveys, we reduced the Orthoptera records to the presence/absence data at the site level and only kept species detected in at least 5\% of sampling sites. We pooled some species' data in a species complex when species confusion was possible. Thus, we aggregated data for species in the genuses *Yersinella*, *Calliptamus*, *Podisma* and in the complex *Chorthippus biggutulus-brunneus-mollis-daimei*. We fitted the three methods to the 24 species kept after data processing.

@@

## Data availability
The data and R scripts used to produce this manuscript are available on GitHub.
